---
title: "Covstretch Function Summary"
author: "Leonardo Wang"
output: pdf_document
header-includes: 
  - \usepackage{amsmath}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(tidyverse)
library(ggradar)
library(janitor)
```

# Introduction

We are interested in three general vaccination strategies:

 - Full dose for anyone 
 - A single optimal fractional dose for anyone
 - Optimal fractional dose for each 10-year age group
 
We will compare their performances under different settings, namely the combinations of the options below: 

 - q: length of vaccination campaign in days for dynamic problems, total dose availability as a fraction of the population for static problems
 - objective: "D" (deaths) or "cumI" (cumulative infections), it is a fraction of the population. It is the performance metric
 - dose_response: "covid_default" or "flu_default", it is vaccine efficacy as a function of dose fraction
 - homogen_mixing: True or False, it is whether people from different age groups make contact with others in homogeneous or heterogeneous frequency. 
 - static: True or False as an indicator of static or dynamic optimization
 - pdeath: "ifr_hic" or "ifr_lic", it is the mortality risk of each age group in high-income or low-income countries
 - scenario: "pars_le_slow" or "pars_le_fast", it essentially indicates how fast the virus spreads upon contact
 - recurring: True or False as an indicator of whether loss of immunity and re-vaccination are considered
 
The code for generating all combinations and apply the three general strategies above can be found in **optimisation-epi/nlopt_al_cases.R**. Run **project_setup.R** before anything. 

The results are stored in **results/results_full_dose.Rdata**, **results/results_frac_uni.Rdata**, and **results/results_frac_age.Rdata**, corresponding to the three general strategies. 

```{r}
load("results/results_all.Rdata")
```

# Comparisons of Objective Values under Different Settings



```{r}
results_all %>% 
  filter(static == T, objective == "D", scenario == "pars_le_slow") %>% 
  mutate(setting = paste("q = ", q, 
                         ", dose response = ", dose_response, 
                        ", homogen mixing = ", homogen_mixing,
                        ", mortality risk = ", pdeath,
                        ", recurring = ", recurring, 
                        sep = "")) %>% 
  radar_plot()
```
all_dynamic <- results_all %>% 
  filter(objective == "D", scenario == "pars_le_slow") %>% 
  mutate(setting = paste("q = ", q, 
                         ", objective = ", objective, 
                         ", dose response = ", dose_response, 
                         ", homogen mixing = ", homogen_mixing, 
                         ", static = ", static,
                         ", mortality risk = ", pdeath, 
                         ", scenario = ", scenario, 
                         ", recurring = ", recurring, 
                         sep = "")) %>% 
  select(setting, full_dose, frac_uni, frac_age) %>% 
  mutate(across(c(full_dose, frac_uni), ~map_dbl(., 1))) %>%
  pivot_longer(cols = c(full_dose, frac_uni, frac_age), 
               names_to = "group", 
               values_to = "value") %>% 
  pivot_wider(names_from = setting, values_from = value)
```

  
```{r}
lcols <- c("#D55E00", "#0072B2", "#009E73")
all_dynamic %>% 
  ggradar(background.circle.colour = "white",
          axis.label.size = 1,
          grid.label.size = 5,
          group.point.size = 1,
          legend.text.size = 5,
          group.line.width = 0.5,
          grid.line.width = 0.1,
          grid.min = 0, 
          grid.max = 0.005 / 4,
          grid.mid = 0.0025 / 4,
          group.colours = lcols) + 
  theme_bw(base_size = 5) +
  theme(legend.position = "bottom")
```


# Optimal age-group-specific dose fraction

```{r}
plot_frac_age("static", F, 
              "pdeath", "ifr_hic",
              "homogen_mixing", F, 
              "scenario", "pars_le_slow", 
              "q", "objective")

plot_frac_age("static", F, 
              "objective", "D",
              "pdeath", "ifr_hic", 
              "scenario", "pars_le_fast", 
              "q", "homogen_mixing")

plot_frac_age("static", F, 
              "objective", "D",
              "homogen_mixing", F, 
              "scenario", "pars_le_slow", 
              "q", "pdeath")
```

\newpage

This note will summarize main functions for simulation and optimization from the **covstretch** repository that can potentially be utilized by the new theoretical paper on optimal doze stretching. 

# Epidemiological Model

We begin with the fundamental epidemiological model. The associated system of differential equations is defined originally in **R/ode_2vaccines_v2.R**, now moved to **review/setup/epi_models.R**, and can be formally written as below. The model contains two vaccine types with different efficacies. Since we are only considering the case of fractional dose, we ignore the second vaccine type in reality. 

\begin{align}
\dot{S}_i(t) &= - \lambda_i(t) S_i(t) + \phi_i R_i(t) - (v_i^1(t) \delta_t^1 - v_i^2(t) \delta_t^2) \frac{S_i(t)}{S_i(t) + R_i(t)} \\
\dot{E}_i(t) &= \lambda_i(t) (S_i(t) + N_i^1(t) + N_i^2(t)) - \gamma_i^{EI} E_i(t) \\
\dot{I}_i(t) &= \gamma_i^{EI} E_i(t) - \gamma_i^{IRD} I_i(t) \\
\dot{R}_i(t) &= (1 - p_i) \gamma_i^{IRD} I_i(t) - \phi_i R_i(t) - (v_i^1 \delta_i^1 + v_i^2 \delta_i^2) \frac{R_i(t)}{S_i(t) + R_i(t)} \\
\dot{D}_i(t) &= p_i \gamma_i^{IRD} I_i(t) \\
\dot{V}_i^1(t) &= v_i^1(t) \delta_i^1 \frac{e_i^1 S_i(t) + R_i(t)}{S_i(t) + R_i(t)} - \kappa_i^1 V_i^1(t) \\
\dot{N}_i^1(t) &= v_i^1(t) \delta_i^1 \frac{(1 - e_i^1) S_i(t)}{S_i(t) +  R_i(t)} + \kappa_i^1 V_i^1(t) - \lambda_i(t) N_i^1(t) \\
\dot{V}_i^2(t) &= v_i^2(t) \delta_i^2 \frac{e_i^2 S_i(t) + R_i(t)}{S_i(t) + R_i(t)} - \kappa_i^2 V_i^2(t) \\
\dot{N}_i^2(t) &= v_i^2(t) \delta_i^2 \frac{(1 - e_i^2) S_i(t)}{S_i(t) + R_i(t)} + \kappa_i^2 V_i^2(t) - \lambda_i(t) N_i^2(t) 
\end{align}

where the definitions can be found in the appendix to the PNAS paper. Note that $\phi_i$ denotes loss of immunity. 

# Parameters

Parameters are defined in **review/parameters_scenario.R** and can be conveniently accessed by the function **grab_2v_parms**. The function **apap_2v** for age prioritization adjustment is stored in **review/parameters_adj_age_prioritization.R**. 

* $\lambda_i(t) = q \sum_{j=1}^G c(i,j) I_j(t)$: rate of new infection, $c(i,j)$ denotes the number of contacts made by an individual in cohort $i$ with another in cohort $j$, $q$ is adjusted match the reproductive number of virus $\mathcal{R}$.
The contact matrix $C$ is contained in **review/setup/default_input.Rdata**. $q$ is 

* $\gamma_i^{EI}$: hazard rate of moving from exposed to infected, set to 0.2. 
* $\gamma_i^{IRD}$: hazard rate of moving from infected to recovered or dead, set to 0.2.
* $p_i$: mortality risk
* $e_i^1$: efficacy of fractional dose. It is by default 0.95. Later in **objective-function.R**, it will be according to a response function of dose fraction. 
* $e_i^2$: efficacy of the second vaccine, set to 0. 
* $v_i$: eligibility of vaccination
* $\delta_i$: rate of vaccination
* $\phi_i$: loss of immunity, currently 0 for all groups

# Initial Values

# Dynamic Optimization

## Single Simulation

The **single_run** function, stored in **review/simulation/single_run.R**, will run the epidemiological model for 360 days. As the real work horse, it takes in parameters and initial values, and outputs the matrix $y$, rows of which are days and columns of which are compartments in the model. 

## Objective Function

Note the **main_matrics** function























