---
title: "Counterfactual for FDF: UK and US"
author: "WW"
date: "08 March 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
as.percent <- function(x, d=2) paste0(round(100*x, d), "%")
```

```{r, include=FALSE}
library(readr)
library(tidyverse)

# UNITED STATES
# Values extracted from CDC source on 24 February 2021
# https://covid.cdc.gov/covid-data-tracker/#vaccination-demographic

# ENGLAND:
# Excel spreadsheet from NHS, extracted from 
# https://www.england.nhs.uk/statistics/statistical-work-areas/covid-19-vaccinations/


# If there is no transmission, then
# Mortality risk = (daily infection risk) x IFR x (length of time unprotected)

load("../data/default_inputs.Rdata")
us_pop <- as.numeric(pbc_spread["840",])
uk_pop <- as.numeric(pbc_spread["826",])
# https://github.com/1DaySooner/RiskModel/blob/master/ifr-model/bayesian_ifr_model.pdf
ifr <- exp(1.12*(1:9 - 3) - 8.80)
us_ifr <- sum(ifr*us_pop)/sum(us_pop)
uk_ifr <- sum(ifr*uk_pop)/sum(uk_pop)

uk_ifr_v <- sum(c(0,0,rep(.44, 5), .34, .2)*ifr)
us_ifr_v <- sum(ifr*c(0, 0.1, 7, 10, 10.5, 12.2, 19.05, 27.1, 14.05)/100)

ifr_all <- .01
efficacy1 <- .8 #efficacy following 1st dose
efficacy2 <- .95 #efficacy following 2nd dose

ifr_v <- c("United Kingdom" = uk_ifr_v, 
           "United States" = us_ifr_v)

# Estimated number of people who had Covid at different points in time
df <- readRDS("owin-data-formatted.rds") %>% 
  filter(location %in% c("United Kingdom", "United States")) %>% 
  filter(date <= "2021-03-08") %>%
  select(location, date, new_deaths_smoothed_per_million, new_cases_smoothed_per_million,
         people_vaccinated_per_hundred, people_fully_vaccinated_per_hundred) %>%
  # left_join(share_people_vaccinated_covid %>% rename(location = Entity, date = Date), by = c("location", "date")) %>%
  # left_join(people_fully_vaccinated_covid %>% rename(location = Entity, date = Date), by = c("location", "date")) %>%
  # select(-Code.x, -Code.y) %>% 
  group_by(location) %>%
  # Count number of infections to date
  mutate(new_infections_per_milion = lead(new_deaths_smoothed_per_million, 30)/uk_ifr) %>% 
  mutate(case_to_inf = new_infections_per_milion[date == "2021-01-24"]/new_cases_smoothed_per_million[date == "2021-01-24"]) %>%
  mutate(new_infections_per_milion = ifelse(is.na(new_infections_per_milion), 
                                            case_to_inf*new_cases_smoothed_per_million, 
                                            new_infections_per_milion)) %>%
  # mutate(new_infections_per_milion = zoo::na.locf(new_infections_per_milion, na.rm=F)) %>% 
  mutate(infected_to_date_per_milion = cumsum(new_infections_per_milion)) %>% 
  # linear approximation for fraction vaccinated:
  mutate(people_vaccinated_per_hundred = zoo::na.approx(people_vaccinated_per_hundred, na.rm=FALSE)) %>%
  mutate(people_vaccinated_per_hundred = 
           ifelse(is.na(people_vaccinated_per_hundred), 0, people_vaccinated_per_hundred)) %>%
  mutate(people_fully_vaccinated_per_hundred = zoo::na.approx(people_fully_vaccinated_per_hundred, na.rm=FALSE)) %>%
  mutate(people_fully_vaccinated_per_hundred = 
           ifelse(is.na(people_fully_vaccinated_per_hundred), 0, people_fully_vaccinated_per_hundred)) %>%
  # Susceptible proportion
  mutate(susceptible_proportion = 1 - 
           efficacy1*(people_vaccinated_per_hundred-people_fully_vaccinated_per_hundred)/100 -
           efficacy2*(people_fully_vaccinated_per_hundred)/100 -
           infected_to_date_per_milion/1e06) %>%
  mutate(new_1st_doses = people_vaccinated_per_hundred/100 - lag(people_vaccinated_per_hundred, 1, default = 0)/100) %>%
  mutate(new_2nd_doses = people_fully_vaccinated_per_hundred/100 - lag(people_fully_vaccinated_per_hundred, 1, default = 0)/100) %>%
  mutate(infection_risk_from_t = (max(infected_to_date_per_milion, na.rm=T) - infected_to_date_per_milion) / 1e06) %>%
  mutate(deaths_averted_by_vaccinations_per_milion = 
           1e06*efficacy1*new_1st_doses*ifr_v[location]*infection_risk_from_t,
         deaths_averted_by_2nd_doses_per_milion = 
           1e06*(efficacy2 - efficacy1)*new_2nd_doses*ifr_v[location]*infection_risk_from_t,
         deaths_averted_by_2nd_doses_ctfl_per_milion = 
           1e06*efficacy1*new_2nd_doses*ifr_v[location]*infection_risk_from_t)


```

```{r}
df_doses <- summarise(df, first = sum(new_1st_doses), second = sum(new_2nd_doses))
# $\dfrac{0.133+0.059 \cdot 0.8}{(0.133-0.059) \cdot 0.8 + 0.95 \cdot 0.059} = 1.33$
ctfl <- (df_doses$first[2]+df_doses$second[2])*efficacy1
basec <- (df_doses$first[2]-df_doses$second[2])*efficacy1 + df_doses$second[2]*efficacy2

```

Consider a simple calculation of how many lives would be saved if the second dose vaccinations were replaced with first doses, without any need to expand capacity. 

For now let us ignore epidemiological externalities and consider reduction in individual risk only. As of `r max(df$date)`, in the UK `r as.percent(df_doses$first[1])` people received first dose and `r as.percent(df_doses$second[1])` received both doses of the vaccine.^[We use data on numbers of: deaths, confirmed cases, vaccinated with first dose, fully vaccinated from Our World In Data. 
] In the US `r as.percent(df_doses$first[2])` people received first dose and `r as.percent(df_doses$second[2])` were fully vaccinated. If we assumed that the first dose of the vaccine reduces mortality risk following infection by $e_1=80\%$ and the second doses reduces it by $e_2=95\%$, then, using basic algebra, the overall number of lives saved in the US by replacing second doses by additional first doses would be `r round(ctfl/basec, 2)` times higher. 

However, this simple calculation cannot tell us how many lives were saved. It also ignores the time dimension. The reduction in mortality risk for someone vaccinated at time $t$ is proportional to, one, risk of death following infection and, two, the risk of infection between time $t$ and now. Thus early vaccinations are more important since they cover the longer time periods. Therefore we repeat the counterfactual calculation by defining infection risk in each day, $i(t)$, as well as vaccinations with first and second doses in each day ($v_1(t)$ and $v_2(t)$, respectively) and then summing over all days since start of mass vaccination programs. The number of deaths averted in the observed scenario is

$$
b = \sum_t [v_1(t)e_1 + v_2(t)(e_2-e_1)]i(t)\text{IFR}_v,
$$

while the number of deaths averted under the counterfactual is 

$$
b^\ast = \sum_t (v_1(t) + v_2(t))e_1i(t)\text{IFR}_v .
$$
We define $\text{IFR}_v$ as follows: since the vaccines have been prioritised for the elderly, the infection fatality risk (IFR) for a typical vaccinated patient is higher than the average IFR in the population. However, we have to account for the fact that many of the early doses are given to health care workers and some of the other key workers. By late February 2021, in the UK around 55% of the vaccines went to people over 70 and over 95% of that age group has been vaccinated. In the US, however, while 55% of vacines went to people over 65, close to 30% went to people younger than 50.^[For NHS data see https://www.england.nhs.uk/statistics/statistical-work-areas/covid-19-vaccinations/ and for the US see https://covid.cdc.gov/covid-data-tracker/#vaccination-demographic] We calculated $\text{IFR}_v$ as an approximate weighted mean of age-specific infection mortality risks, using a meta-analysis estimate in Manheim _et al._, 2021.^[For example, since the IFR in 80+ year olds is 12% and in 70-80 it's 4%, then distributing 50% of doses to 70-79 and 50% to 80+ year olds would have $\text{IFR}_v$ of 8%.] 

Applying this $\text{IFR}_v$ approach to real-world distributions of vaccine distribution, for UK we obtained `r as.percent(uk_ifr_v, 1)` and for the US `r as.percent(us_ifr_v, 1)`, a remarkable difference. In other words, __despite delivering twice the number of doses (and "running out" of highest risk individuals to vaccinate), a single dose of vaccine in the UK was still used 50% more effectively than in the US.__ (It should be noted, however, that the UK has a slightly older population than the US.)


```{r}
fo <- function(x) format(x, digits = 0, big.mark = ",", scientific = F)

df_result <- df %>%
  summarise(
    d1 = sum(new_1st_doses),
    d2 = sum(new_2nd_doses),
    v1 = sum(deaths_averted_by_vaccinations_per_milion, na.rm=T),
    b1 = sum(deaths_averted_by_2nd_doses_per_milion, na.rm=T),
    b2 = sum(deaths_averted_by_2nd_doses_ctfl_per_milion, na.rm=T)
    ) %>%
  mutate(pop = c(66.65, 328.2)) %>%
  mutate(v1_t = pop*v1, b1_t = pop*b1, b2_t = pop*b2) %>%
  mutate(total_b = v1_t+b1_t, total_b_ast = v1_t+b2_t, rr = (v1+b2)/(v1+b1))
```

Next, we calculated the number of lives saved by observed vaccination policy $b$ and the no-second-doses counterfactual $b^*$. We found that in the UK `r fo(df_result$total_b[1])` deaths were averted by vaccination campaign up to `r max(df$date)`, compared to `r fo(df_result$total_b[2])` in the US. That corresponds to, respectively, `r fo(df_result$b1[1] + df_result$v1[1])` and `r fo(df_result$b1[2] + df_result$v1[2])` deaths averted per million people. It's important to note that this reduction occured over a very short period of time, with an average vaccinated person being protected only for a few weeks -- both of these numbers will of course increase over time. As we mentioned above, the almost 3-fold difference is not only due to higher capacity in the UK but also due to better targeting of available doses.

The reductions that could be achieved by replacing the second dose visits with first doses (thus requiring no additional capacity) are additional `r fo(df_result$total_b_ast[1] - df_result$total_b[1])` deaths averted in the UK and `r fo(df_result$total_b_ast[2] - df_result$total_b[2])` in the US. That corresponds to `r as.percent(df_result$rr[2] - 1, 0)` additional deaths that could have been avoided in the US, even if we assume a drop in efficacy to 80% and ignore the effect of vaccine on breaking the transmission.




```{r, include = FALSE}
df %>% ggplot(aes(x = date, y = new_infections_per_milion, color = location)) + geom_line()
df %>% ggplot(aes(x = date, y = new_deaths_smoothed_per_million, color = location)) + geom_line()

df %>% 
  ggplot(aes(x = date, y = infection_risk_from_t, color = location)) + geom_line()
```

